# Thinking Like a Bayesian

```{r}
#| label: setup-ch03
#| include: false
source("../R/setup.R")
```

## Overview

This chapter builds the conceptual foundation for everything that follows.
Before touching Stan again, you need a working mental model of three ideas:
the **likelihood**, the **prior**, and the **posterior** — and how they
combine through Bayes' theorem. You also need an intuition for what
Markov chain Monte Carlo (MCMC) is doing when Stan runs.

None of this requires measure theory or a probability textbook. The goal is
a set of pictures and metaphors that make the Stan output in later chapters
feel *interpretable*, not mysterious.

## Learning objectives

By the end of this chapter you will be able to:

- State Bayes' theorem in words and explain what each term means.
- Distinguish the likelihood, the prior, and the posterior, and describe
  how each one changes as data accumulate.
- Explain intuitively what MCMC does and why we use samples rather than
  closed-form solutions.
- Describe three types of prior (flat, weakly informative, strongly
  informative) and give a reason to prefer each.

---

## A running example: soil nitrogen

Throughout this chapter we use a concrete, agronomic example. Suppose you
are measuring total nitrogen concentration (g kg⁻¹) in topsoil samples
from a fallow plot in the forest-savanna transition zone of Côte d'Ivoire.
You have a prior belief about nitrogen levels from the literature, you
collect a small number of samples, and you want to update your knowledge.

This is the Bayesian problem in miniature: combining what you knew before
with what the data show.

---

## Bayes' theorem

Bayes' theorem is a statement about conditional probability. In the context
of statistical modelling it reads:

$$
\underbrace{p(\theta \mid y)}_{\text{posterior}}
\;\propto\;
\underbrace{p(y \mid \theta)}_{\text{likelihood}}
\times
\underbrace{p(\theta)}_{\text{prior}}
$$

where $\theta$ is the parameter (or vector of parameters) you want to learn,
and $y$ is the data you observed. The $\propto$ symbol means
"proportional to" — we drop the normalising constant $p(y)$ (the
*marginal likelihood*) because Stan never needs to compute it.

Let us unpack each term.

### The prior: $p(\theta)$

The prior encodes what you believe about $\theta$ *before seeing the data*.
It is a probability distribution over all plausible values of $\theta$.

For our nitrogen example we can draw on measured values from the literature.
In mid-west Côte d'Ivoire, @yao2010 reported topsoil (0–10 cm) total
nitrogen concentrations ranging from roughly 0.7 g kg⁻¹ in degraded
cropland to around 1.6 g kg⁻¹ under secondary forest across several land
use types. In the forest-savanna transition zone of central Côte d'Ivoire,
@ilboudo2022 measured total nitrogen across 160 sites (0–20 cm) under
different land uses, with values consistent with that range under natural
vegetation. A comparable study in the forest-savanna transition of Cameroon
found topsoil N of 1.6–2.5 g kg⁻¹ in native and fallow lands at 0–10 cm
depth [@djouama2024]. Taken together, these studies suggest that in
undisturbed or fallow plots of the West African forest-savanna mosaic, mean
topsoil N values of approximately 1.0–1.8 g kg⁻¹ are plausible, with
cropped and degraded soils falling below 1.0 g kg⁻¹.

For a fallow plot in this system, a reasonable prior might therefore be:

$$
\mu \sim \text{Normal}(1.2, 0.4)
$$

This places 95% prior probability between 0.4 and 2.0 g kg⁻¹, ruling out
values below zero or above 3 g kg⁻¹ as implausible, while remaining
consistent with the range documented in the literature. The centre of 1.2
is conservative — appropriate for a mixed fallow-forest landscape rather
than claiming certainty about the exact habitat type.

::: {.callout-note}
**A prior is not a guess.** It is a *distribution* — a full description of
your uncertainty before the experiment. A prior centred at 1.2 with a wide
standard deviation expresses much less certainty than one centred at 1.2 with
a narrow standard deviation. Both say the same thing about the centre; they
say very different things about how much you trust it.
:::

### The likelihood: $p(y \mid \theta)$

The likelihood measures *how probable the data are, given a particular value
of $\theta$*. It is not a probability distribution over $\theta$ — it is a
function of $\theta$ that scores how well each candidate value of $\theta$
explains the data you actually observed.

If we assume nitrogen measurements are approximately normally distributed,
the likelihood for a single observation $y_i$ given mean $\mu$ and standard
deviation $\sigma$ is:

$$
p(y_i \mid \mu, \sigma)
= \frac{1}{\sigma\sqrt{2\pi}}
  \exp\!\left(-\frac{(y_i - \mu)^2}{2\sigma^2}\right)
$$

For $N$ independent observations, the joint likelihood is the product:

$$
p(y \mid \mu, \sigma) = \prod_{i=1}^{N} p(y_i \mid \mu, \sigma)
$$

In practice — and in Stan — we work with the *log* likelihood because
products of small numbers underflow numerically.

### The posterior: $p(\theta \mid y)$

The posterior is what you believe about $\theta$ *after seeing the data*. It
combines the prior (what you knew) with the likelihood (what the data say)
into an updated distribution.

The posterior is the answer to your inferential question. Every credible
interval, every probability statement, every prediction in this handbook
comes from the posterior.

---

## Updating in pictures

The best way to feel Bayes' theorem is to watch the posterior shift as data
arrive. @fig-updating shows three snapshots: the prior alone, the likelihood
from five observations, and the resulting posterior.

```{r}
#| label: fig-updating
#| fig-cap: >
#|   Bayesian updating for soil nitrogen. The prior (dashed) encodes
#|   pre-experiment knowledge. The likelihood (dotted) summarises the
#|   information in five measurements. The posterior (solid) is their
#|   product, renormalised. As data accumulate, the posterior narrows and
#|   shifts toward the data.
#| fig-width: 7
#| fig-height: 3.8

library(ggplot2)

# Prior: Normal(1.2, 0.4)
prior_mu  <- 1.2
prior_sd  <- 0.4

# Observed data (5 nitrogen measurements, g/kg)
set.seed(42)
y_obs <- rnorm(5, mean = 1.6, sd = 0.3)
n     <- length(y_obs)

# Known sigma for simplicity (we will relax this in Chapter 4)
sigma <- 0.35

# Conjugate posterior for Normal-Normal model
# posterior mean and sd (known sigma case)
post_var <- 1 / (1/prior_sd^2 + n/sigma^2)
post_sd  <- sqrt(post_var)
post_mu  <- post_var * (prior_mu/prior_sd^2 + sum(y_obs)/sigma^2)

# Likelihood: proportional to Normal(ybar, sigma/sqrt(n))
lik_mu <- mean(y_obs)
lik_sd <- sigma / sqrt(n)

mu_grid <- seq(0.2, 2.8, length.out = 500)

df <- data.frame(
  mu       = rep(mu_grid, 3),
  density  = c(
    dnorm(mu_grid, prior_mu, prior_sd),
    dnorm(mu_grid, lik_mu,   lik_sd),
    dnorm(mu_grid, post_mu,  post_sd)
  ),
  component = rep(
    c("Prior  p(μ)",
      "Likelihood  p(y|μ)",
      "Posterior  p(μ|y)"),
    each = length(mu_grid)
  )
)

df$component <- factor(
  df$component,
  levels = c("Prior  p(μ)", "Likelihood  p(y|μ)", "Posterior  p(μ|y)")
)

cols <- c(
  "Prior  p(μ)"          = "#aec6c8",
  "Likelihood  p(y|μ)"   = "#5c8a52",
  "Posterior  p(μ|y)"    = "#2e7d6a"
)
ltys <- c(
  "Prior  p(μ)"          = "dashed",
  "Likelihood  p(y|μ)"   = "dotted",
  "Posterior  p(μ|y)"    = "solid"
)

ggplot(df, aes(mu, density, colour = component, linetype = component)) +
  geom_line(linewidth = 1.0) +
  geom_vline(xintercept = mean(y_obs), colour = "#5c8a52",
             linetype = "dotted", alpha = 0.4) +
  annotate("text", x = mean(y_obs) + 0.06, y = 2.55,
           label = paste0("ȳ = ", round(mean(y_obs), 2)),
           colour = "#5c8a52", size = 3.2, hjust = 0) +
  scale_colour_manual(values = cols, name = NULL) +
  scale_linetype_manual(values = ltys, name = NULL) +
  scale_x_continuous(breaks = seq(0.4, 2.8, by = 0.4)) +
  labs(
    x = "μ  (mean soil nitrogen, g kg⁻¹)",
    y = "Density"
  ) +
  theme(
    legend.position  = c(0.78, 0.82),
    legend.text      = element_text(size = 9),
    legend.key.width = unit(1.8, "cm")
  )
```

Notice three things in @fig-updating:

1. The **posterior lies between the prior and the likelihood**. It is a
   weighted average, with the weights determined by relative precision.
2. The posterior is **narrower** than either the prior or the likelihood
   alone — combining two sources of information reduces uncertainty.
3. When the data ($\bar{y} \approx$ `r round(mean(y_obs), 2)`) disagree with
   the prior centre (1.2), the posterior is pulled toward the data, because
   five measurements carry more information than a vague prior.

### What happens as data accumulate?

@fig-sequential shows the posterior after 1, 5, and 20 observations from the
same generating process. As $N$ grows, the prior is overwhelmed and the
posterior converges on the data-generating mean.

```{r}
#| label: fig-sequential
#| fig-cap: >
#|   Sequential Bayesian updating. Each new observation sharpens the
#|   posterior. With 20 observations the prior is nearly irrelevant and
#|   the posterior is tightly concentrated around the sample mean.
#| fig-width: 7
#| fig-height: 3.5

set.seed(7)
y_all <- rnorm(20, mean = 1.6, sd = 0.35)
ns    <- c(1, 5, 20)

posterior_params <- function(n_obs, y, prior_mu, prior_sd, sigma) {
  pv  <- 1 / (1/prior_sd^2 + n_obs/sigma^2)
  psd <- sqrt(pv)
  pmu <- pv * (prior_mu/prior_sd^2 + sum(y[seq_len(n_obs)])/sigma^2)
  list(mu = pmu, sd = psd)
}

rows <- lapply(ns, function(n_obs) {
  pp <- posterior_params(n_obs, y_all, prior_mu, prior_sd, sigma)
  data.frame(
    mu      = mu_grid,
    density = dnorm(mu_grid, pp$mu, pp$sd),
    n_label = paste0("N = ", n_obs)
  )
})

df_seq <- do.call(rbind, rows)
df_seq$n_label <- factor(df_seq$n_label,
                         levels = paste0("N = ", ns))

prior_df <- data.frame(
  mu = mu_grid,
  density = dnorm(mu_grid, prior_mu, prior_sd),
  n_label = "Prior"
)
prior_df$n_label <- factor(prior_df$n_label)

seq_cols <- c(
  "Prior" = "#cccccc",
  "N = 1" = "#aec6c8",
  "N = 5" = "#5c8a52",
  "N = 20" = "#2e7d6a"
)

ggplot(
  rbind(
    transform(prior_df, n_label = "Prior"),
    df_seq
  ),
  aes(mu, density,
      colour = factor(n_label,
                      levels = c("Prior", "N = 1", "N = 5", "N = 20")),
      linewidth = factor(n_label,
                         levels = c("Prior", "N = 1", "N = 5", "N = 20")))
) +
  geom_line() +
  scale_colour_manual(values = seq_cols, name = "Posterior after") +
  scale_linewidth_manual(
    values = c("Prior" = 0.5, "N = 1" = 0.7, "N = 5" = 0.9, "N = 20" = 1.1),
    name = "Posterior after"
  ) +
  scale_x_continuous(breaks = seq(0.4, 2.8, by = 0.4)) +
  labs(x = "μ  (mean soil nitrogen, g kg⁻¹)", y = "Density") +
  theme(legend.position = c(0.82, 0.72))
```

---

## Choosing a prior {#sec-priors}

Perhaps no aspect of Bayesian analysis generates more anxiety — and more
unnecessary debate — than prior choice. Here is a practical framework.

### Flat (uninformative) priors

A flat prior assigns equal density to all values of $\theta$:

$$
p(\theta) \propto 1
$$

This sounds appealing — "let the data speak" — but flat priors are often
problematic in practice:

- On bounded parameters (like $\sigma > 0$), a flat prior puts infinite mass
  near zero, which can cause sampling difficulties.
- In hierarchical models, flat priors on variance components can lead to
  improper posteriors.
- "Flat" is not scale-invariant: a flat prior on $\sigma$ is not flat on
  $\log \sigma$.

Flat priors are rarely the right choice. Use them only when you genuinely
have no information *and* you have verified the posterior is proper.

### Weakly informative priors

A weakly informative prior rules out values that are physically or
biologically implausible, without strongly favouring any particular value in
the plausible range:

$$
\mu \sim \text{Normal}(0, 10) \qquad \sigma \sim \text{Exponential}(1)
$$

These are the workhorses of applied Bayesian modelling and the defaults in
this handbook. The `Normal(0, 10)` prior for a mean says: "I don't think the
mean is outside the range (−30, 30), but within that range I have little
preference." The `Exponential(1)` prior for a standard deviation says:
"Values near zero are unlikely; moderate values are plausible; very large
values are increasingly surprising."

::: {.callout-tip}
**A useful heuristic:** before fitting, ask yourself "what is the largest
effect I could plausibly see in this system?" Set your prior so that this
value lies somewhere in the 95% interval. If the largest plausible nitrogen
difference between treatments is 1.5 g kg⁻¹, a `Normal(0, 1)` prior on
a treatment effect (95% interval: −1.96, +1.96) is weakly informative.
:::

### Strongly informative priors

When you have reliable external knowledge — a well-studied system, a
meta-analysis, replicated prior experiments — you can use an informative prior
to incorporate that knowledge explicitly:

$$
\mu \sim \text{Normal}(1.2, 0.2)
$$

Strongly informative priors improve estimates when data are scarce. They also
make your assumptions transparent: instead of hiding prior knowledge in a
subjective choice of model structure, you state it as a distribution that
reviewers can inspect and challenge.

@fig-prior-types shows the three prior families for a mean parameter.

```{r}
#| label: fig-prior-types
#| fig-cap: >
#|   Three prior types for a location parameter. The flat prior (dashed,
#|   truncated for display) gives equal weight to implausible values.
#|   The weakly informative prior rules out extremes without strong
#|   commitment. The strongly informative prior encodes precise external
#|   knowledge.
#| fig-width: 7
#| fig-height: 3.5

mu_g <- seq(-4, 6, length.out = 600)

df_priors <- data.frame(
  mu      = rep(mu_g, 3),
  density = c(
    dunif(mu_g, -10, 10) * 3,          # flat (scaled for display)
    dnorm(mu_g, 0, 2),                  # weakly informative
    dnorm(mu_g, 1.2, 0.3)              # strongly informative
  ),
  type = rep(
    c("Flat  Uniform(−10, 10)",
      "Weakly informative  Normal(0, 2)",
      "Strongly informative  Normal(1.2, 0.3)"),
    each = length(mu_g)
  )
)

df_priors$type <- factor(
  df_priors$type,
  levels = c(
    "Flat  Uniform(−10, 10)",
    "Weakly informative  Normal(0, 2)",
    "Strongly informative  Normal(1.2, 0.3)"
  )
)

pcols <- c(
  "Flat  Uniform(−10, 10)"                = "#cccccc",
  "Weakly informative  Normal(0, 2)"       = "#5c8a52",
  "Strongly informative  Normal(1.2, 0.3)" = "#2e7d6a"
)
pltys <- c(
  "Flat  Uniform(−10, 10)"                = "dashed",
  "Weakly informative  Normal(0, 2)"       = "solid",
  "Strongly informative  Normal(1.2, 0.3)" = "solid"
)

ggplot(df_priors, aes(mu, density, colour = type, linetype = type)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = pcols, name = NULL) +
  scale_linetype_manual(values = pltys, name = NULL) +
  coord_cartesian(ylim = c(0, 1.5)) +
  labs(x = "Parameter value (μ)", y = "Prior density") +
  theme(
    legend.position  = c(0.62, 0.80),
    legend.text      = element_text(size = 8.5),
    legend.key.width = unit(1.5, "cm")
  )
```

---

## What MCMC is doing {#sec-mcmc}

For any realistically complex model, the posterior $p(\theta \mid y)$ cannot
be computed analytically — the normalising constant $p(y)$ involves an
integral over all possible values of $\theta$, which is intractable in more
than a few dimensions. Instead, Stan uses **Markov chain Monte Carlo** (MCMC)
to *sample* from the posterior without ever computing it directly.

### The core idea

Imagine the posterior as a landscape, where height represents probability
density. The goal is to draw a map of this landscape by visiting locations
in proportion to their height — spending more time in high-probability
regions, less time in low-probability ones.

MCMC builds a Markov chain — a sequence of parameter values
$\theta^{(1)}, \theta^{(2)}, \ldots, \theta^{(S)}$ — that, after a
*warmup* period, visits each region of the parameter space in proportion
to the posterior density. The resulting collection of draws is a **sample
from the posterior**.

From this sample you can compute any posterior summary:

$$
\mathbb{E}[\theta \mid y] \approx \frac{1}{S} \sum_{s=1}^{S} \theta^{(s)}
\qquad
\text{Var}[\theta \mid y] \approx \frac{1}{S-1}
  \sum_{s=1}^{S}\!\left(\theta^{(s)} - \bar\theta\right)^2
$$

A 90% credible interval is simply the interval from the 5th to the 95th
percentile of $\{\theta^{(s)}\}$.

### Why Stan's MCMC is especially good

Classic MCMC algorithms (Gibbs sampling, used by JAGS and BUGS) propose new
parameter values by taking random steps. In high-dimensional or
strongly-correlated posteriors, most random steps are rejected — the chain
mixes slowly and you need millions of iterations to get reliable estimates.

Stan uses **Hamiltonian Monte Carlo** (HMC), which borrows an analogy from
physics. It treats the negative log posterior as a potential energy surface
and simulates the trajectory of a particle with momentum rolling across that
surface. The particle moves efficiently through curved, high-dimensional
posteriors that would defeat a random-walk sampler [@betancourt2017conceptual].

The specific variant Stan uses — the **No-U-Turn Sampler** (NUTS) — also
automatically tunes the step size and trajectory length during warmup, so
you rarely need to touch these settings manually.

@fig-mcmc-paths illustrates the difference in exploration efficiency
between a slow random-walk chain and an efficient HMC path.

```{r}
#| label: fig-mcmc-paths
#| fig-cap: >
#|   Schematic comparison of random-walk MCMC (left) and Hamiltonian Monte
#|   Carlo (right) on a correlated bivariate posterior. The random walk
#|   takes many small, mostly rejected steps. HMC follows smooth
#|   trajectories that traverse the posterior efficiently.
#| fig-width: 8
#| fig-height: 4

set.seed(17)

# Simulate a correlated bivariate target (elongated ellipse)
Sigma <- matrix(c(1, 0.85, 0.85, 1), 2, 2)
L     <- chol(Sigma)

# Random-walk path (small steps, slow exploration)
rw_path <- matrix(c(0, 0), nrow = 1)
for (i in seq_len(300)) {
  proposal <- tail(rw_path, 1) + rnorm(2, 0, 0.18) %*% L
  log_r    <- -(sum(proposal^2) - sum(tail(rw_path,1)^2)) / 2
  if (log(runif(1)) < log_r) {
    rw_path <- rbind(rw_path, proposal)
  } else {
    rw_path <- rbind(rw_path, tail(rw_path, 1))
  }
}
rw_df <- data.frame(
  x = rw_path[, 1], y = rw_path[, 2],
  step = seq_len(nrow(rw_path)),
  method = "Random-walk MCMC"
)

# HMC-like smooth trajectory (simulate leapfrog arcs)
hmc_pts <- list(c(-1.5, -1.5))
for (i in seq_len(18)) {
  start  <- tail(hmc_pts, 1)[[1]]
  theta  <- runif(1, 0, 2*pi)
  radius <- runif(1, 0.6, 1.4)
  arc    <- seq(theta, theta + runif(1, pi*0.4, pi*0.9), length.out = 12)
  pts    <- lapply(arc, function(a)
    start + c(radius * cos(a), radius * 0.55 * sin(a)) %*% L)
  hmc_pts <- c(hmc_pts, pts)
}
hmc_mat <- do.call(rbind, hmc_pts)
hmc_df  <- data.frame(
  x = hmc_mat[, 1], y = hmc_mat[, 2],
  step = seq_len(nrow(hmc_mat)),
  method = "Hamiltonian Monte Carlo"
)

# Contour data
grid  <- expand.grid(x = seq(-3, 3, length.out = 80),
                     y = seq(-3, 3, length.out = 80))
z_val <- with(grid,
  exp(-0.5 * (x^2 - 2*0.85*x*y + y^2) / (1 - 0.85^2)))
grid$z <- z_val

both <- rbind(rw_df, hmc_df)

ggplot() +
  geom_contour(data = grid, aes(x, y, z = z),
               colour = "grey80", linewidth = 0.35, bins = 7) +
  geom_path(data = both,
            aes(x, y, colour = method, alpha = step),
            linewidth = 0.55, show.legend = FALSE) +
  geom_point(data = both[seq(1, nrow(both), by = 6), ],
             aes(x, y, colour = method),
             size = 1.0, alpha = 0.6) +
  scale_colour_manual(
    values = c("Random-walk MCMC" = "#c0392b",
               "Hamiltonian Monte Carlo" = "#2e7d6a"),
    name = NULL
  ) +
  scale_alpha_continuous(range = c(0.2, 0.9), guide = "none") +
  facet_wrap(~method) +
  labs(x = expression(theta[1]), y = expression(theta[2])) +
  theme(
    strip.text      = element_text(face = "bold", size = 10),
    legend.position = "none"
  )
```

### Warmup and sampling phases

Stan's sampler runs in two phases:

**Warmup** (`iter_warmup`, default 1000): the chain explores the posterior
and automatically tunes its step size and mass matrix. *Warmup draws are
discarded* — they are not part of your posterior sample.

**Sampling** (`iter_sampling`, default 1000): the tuned chain draws samples
that are retained for inference.

With four chains, the default settings give $4 \times 1000 = 4000$ posterior
draws — usually more than enough for well-behaved models.

### Multiple chains

Stan runs several chains (typically four) starting from different random
initial values. If all chains converge to the same region of the posterior,
that is strong evidence of a valid result. The $\hat{R}$ statistic measures
this agreement: values close to 1.00 indicate that between-chain variance
equals within-chain variance, as expected when all chains explore the same
distribution [@vehtari2021rank].

---

## From posterior draws to answers

Once you have posterior draws, every inferential question is a calculation:

```{r}
#| label: tbl-posterior-calc
#| echo: false
#| tbl-cap: "Translating questions into posterior calculations."

df_calc <- data.frame(
  Question = c(
    "What is the most likely value of μ?",
    "How uncertain am I about μ?",
    "What is the 90% credible interval for μ?",
    "P(μ > 1.5)?",
    "What will a new observation look like?"
  ),
  Calculation = c(
    "median(draws$mu)",
    "sd(draws$mu)",
    "quantile(draws$mu, c(0.05, 0.95))",
    "mean(draws$mu > 1.5)",
    "draws$y_rep  (from generated quantities)"
  )
)

knitr::kable(df_calc, col.names = c("Inferential question", "R calculation on draws"))
```

This is the practical payoff of Bayesian inference: answers to questions
that matter for decision-making, expressed as probabilities rather than
binary hypothesis tests.

---

## The Bayesian workflow

Before closing, it helps to see how the three concepts — prior, likelihood,
posterior — fit into a complete analysis cycle (@fig-workflow). This cycle
recurs in every chapter of this handbook.

```{r}
#| label: fig-workflow
#| echo: false
#| fig-cap: >
#|   The Bayesian modelling workflow. Arrows show the flow from scientific
#|   question to posterior and back. Prior predictive checks (Chapter 13)
#|   and posterior predictive checks (Chapters 4–12) close the loop.
#| fig-width: 7
#| fig-height: 3.8

library(ggplot2)

nodes <- data.frame(
  x     = c(1,   3,   5,   3,   5),
  y     = c(2.5, 4,   4,   1,   1),
  label = c(
    "Scientific\nquestion",
    "Prior\np(θ)",
    "Likelihood\np(y|θ)",
    "Prior predictive\ncheck",
    "Posterior\np(θ|y)"
  ),
  fill  = c("#e8f4f2", "#aec6c8", "#5c8a52", "#f0ede8", "#2e7d6a"),
  tcol  = c("grey30",  "grey20",  "white",   "grey30",  "white")
)

edges <- data.frame(
  x    = c(1.55, 3,    3.55, 3,   4.55),
  y    = c(2.85, 3.65, 4,    3.65, 1  ),
  xend = c(2.45, 3,    4.45, 3,   5.45),
  yend = c(3.65, 1.35, 4,    1.35, 1  ),
  lbl  = c("encode", "simulate", "×  data", "update", "use")
)

ggplot() +
  geom_segment(data = edges,
               aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
               colour = "grey50", linewidth = 0.7) +
  geom_label(data = nodes,
             aes(x, y, label = label, fill = fill, colour = tcol),
             size = 3.2, fontface = "bold",
             label.padding = unit(0.45, "lines"),
             label.r = unit(0.3, "lines"),
             show.legend = FALSE) +
  scale_fill_identity() +
  scale_colour_identity() +
  annotate("text", x = 1.95, y = 3.38, label = "encode",
           size = 2.8, colour = "grey50", fontface = "italic") +
  annotate("text", x = 3.0,  y = 2.50, label = "prior\npredictive",
           size = 2.8, colour = "grey50", fontface = "italic", hjust = 0.5) +
  annotate("text", x = 4.0,  y = 4.18, label = "× data",
           size = 2.8, colour = "grey50", fontface = "italic") +
  annotate("text", x = 3.0,  y = 2.50-1.15, label = "posterior\nupdating",
           size = 2.8, colour = "grey50", fontface = "italic", hjust = 0.5) +
  coord_cartesian(xlim = c(0.4, 6.0), ylim = c(0.3, 4.8)) +
  theme_void()
```

---

## Summary

- **Bayes' theorem** combines the prior $p(\theta)$ and the likelihood
  $p(y \mid \theta)$ into the posterior $p(\theta \mid y)$. The posterior
  is proportional to their product.
- The **prior** encodes pre-experiment knowledge as a distribution over
  parameters. Weakly informative priors are the practical default: they rule
  out implausible values without dominating the data.
- The **likelihood** scores how probable the observed data are under each
  candidate parameter value. It is a function of $\theta$, not a
  distribution over $\theta$.
- The **posterior** narrows and shifts toward the data as observations
  accumulate. With enough data, the prior is overwhelmed.
- **MCMC** draws samples from the posterior without computing it directly.
  Stan uses Hamiltonian Monte Carlo, which is efficient in high dimensions.
  Multiple chains and the $\hat{R}$ diagnostic provide convergence assurance.
- Every posterior quantity — mean, interval, probability statement,
  prediction — is a simple calculation on the sample of draws.

---

## Exercises

1. **Prior sensitivity.** Using the conjugate normal model from Chapter 2,
   fit the same data three times with: (a) `Normal(0, 10)`, (b)
   `Normal(0, 0.5)`, and (c) `Normal(5, 0.2)` as the prior on `mu`.
   For each fit, extract the posterior mean and 90% credible interval.
   How much does the prior influence the posterior? At what sample size does
   the influence disappear?

2. **Posterior probabilities.** Fit the nitrogen model from this chapter
   (or any normal model). Using the posterior draws, compute:
   (a) $P(\mu > 1.5 \mid y)$, (b) $P(0.8 < \mu < 2.0 \mid y)$, and
   (c) the posterior probability that a new observation will exceed 2.5.
   Write the one-line R code for each.

3. **Prior predictive simulation.** Before fitting, draw 1 000 values of
   $\mu$ from `Normal(1.2, 0.4)` and $\sigma$ from `Exponential(1)`. For
   each pair $(\mu, \sigma)$, draw one observation from
   `Normal(mu, sigma)`. Plot the resulting distribution of synthetic
   observations. Does the prior imply any nitrogen values you would
   consider physically implausible? Adjust the prior and repeat.
   *(We will formalise this as prior predictive checks in Chapter 13.)*

4. **Credible interval vs confidence interval.** A 95% Bayesian credible
   interval $[a, b]$ means: "Given the data and the prior, there is a 95%
   probability that $\theta$ lies in $[a, b]$." A frequentist 95%
   confidence interval means something subtly different. Write one or two
   sentences explaining the difference in plain language. Which
   interpretation is more useful for an agronomist making a management
   decision?

---

## Further reading

- @mcelreath2020rethinking [Ch. 2–3] — the best intuitive treatment of
  Bayes' theorem for life scientists, using grid approximation to build
  visual understanding before MCMC.
- @gelman2013bda [Ch. 1–2] — the authoritative technical treatment; skim
  Ch. 1 for perspective and Ch. 2 for the conjugate normal model worked
  out in full.
- @betancourt2017conceptual — if you want to understand *why* HMC works,
  §2–3 give a geometric intuition without requiring physics background.
- @vehtari2021rank — the definitive reference for $\hat{R}$ and ESS;
  explains why the older $\hat{R}$ was subtly flawed and what the
  rank-normalised version fixes.
