# Setting Up Your Environment

```{r}
#| label: setup-ch02
#| include: false
source("../R/setup.R")
```

## Overview

Before you can write a Stan model, you need a working toolchain: a C++
compiler, CmdStan itself, and the R interface **cmdstanr**. This chapter
walks through every step, shows you how to verify that each piece is working,
and ends with a complete first Stan program that you compile and run yourself.

If you hit a problem, the troubleshooting section at the end of this chapter
covers the most common installation failures by operating system.

## Learning objectives

By the end of this chapter you will be able to:

- Install CmdStan and cmdstanr on your machine and verify the toolchain.
- Describe the structure of a `.stan` file (the six named blocks).
- Compile and sample from a minimal Stan program using `cmdstanr`.
- Read and interpret the summary output of a fitted Stan model.


## The toolchain: what you are installing

Stan models are written in the Stan language, but they are **compiled to C++**
before sampling. This means you need:

1. **A C++ compiler**: the Stan compiler (`stanc`) translates `.stan` =>
   `.cpp`; your system compiler then builds a fast binary.
2. **CmdStan**: the command-line Stan interface; the actual sampler lives
   here.
3. **cmdstanr**: an R package that calls CmdStan from R, passes your data,
   and returns results as tidy R objects.

```{mermaid}
%%| label: fig-toolchain
%%| fig-cap: "The Stan compilation and sampling toolchain."
%%| fig-align: center

flowchart TD
    A("📄 Your .stan file"):::user
    B("🔧 stanc\nStan → C++"):::tool
    C("📝 generated .cpp"):::artifact
    D("⚙️ C++ compiler\ng++ / clang++"):::tool
    E("💾 compiled binary\n.exe / .out"):::artifact
    F("📦 cmdstanr\ncalls the binary,\npasses your data"):::tool
    G("📊 posterior samples\nin R"):::result

    A --> B --> C --> D --> E --> F --> G

    classDef user     fill:#e8f4f2,stroke:#2e7d6a,color:#1a2e28
    classDef tool     fill:#2e7d6a,stroke:#1a4a40,color:#ffffff
    classDef artifact fill:#f0f5f4,stroke:#5c8a52,color:#1a2e28
    classDef result   fill:#5c8a52,stroke:#3a6035,color:#ffffff
```

::: {.callout-note}
**Why cmdstanr, not rstan?**  
Two R interfaces exist for Stan: `rstan` (the original) and `cmdstanr` (the
newer, recommended one). This handbook uses `cmdstanr` throughout because it tracks CmdStan releases closely, has a cleaner API, and avoids the compilation overhead that can make rstan slow to load. If you already use rstan, the Stan code in this handbook is identical, only the R calls differ.
:::


## Step 1: Install a C++ compiler

### Windows

Install **Rtools** for your R version from
<https://cran.r-project.org/bin/windows/Rtools/>. Accept all defaults.
Rtools provides the `g++` compiler that Stan needs.

After installation, verify in R:

```{r}
#| eval: false
pkgbuild::has_build_tools(debug = TRUE)
```

You should see `TRUE`. If not, restart R and try again.

### macOS

You need the **Xcode Command Line Tools**, which provide the `clang++`
compiler:

```bash
xcode-select --install
```

A dialog will appear; click *Install*. After it completes, verify:

```bash
clang++ --version
```

If you see a version string (e.g., `Apple clang version 15.0.0`), you are ready. On Apple Silicon (M1/M2/M3), this is sufficient and no additional steps are needed.

### Linux

Most Linux distributions ship with `g++`. Check:

```bash
g++ --version
```

If the command is not found, install the build tools:

```bash
# Debian / Ubuntu
sudo apt-get install build-essential

# Fedora / RHEL
sudo dnf groupinstall "Development Tools"
```


## Step 2: Install cmdstanr

`cmdstanr` is not yet on CRAN; install it from the Stan R-universe:

```{r}
#| eval: false
install.packages(
  "cmdstanr",
  repos = c("https://stan-dev.r-universe.dev", getOption("repos"))
)
```

Load it and check that R can see it:

```{r}
#| eval: false
library(cmdstanr)
cmdstanr:::check_cmdstan_toolchain()
```

If the toolchain check passes, you will see a message like:

```
The C++ toolchain required for CmdStan is setup properly!
```

If it fails, revisit Step 1 and consult the troubleshooting section below.


## Step 3: Install CmdStan

With cmdstanr loaded, installing CmdStan is a single R call:

```{r}
#| eval: false
install_cmdstan(cores = 2)   # increase cores to speed up compilation
```

This downloads and compiles CmdStan. It takes 5–15 minutes on the first
install. You only need to do this once (and again when you want to upgrade).

::: {.callout-tip}
`install_cmdstan()` stores CmdStan in `~/.cmdstan/` by default. You can
check and change this path with `cmdstan_path()` and the `CMDSTAN`
environment variable in `.Renviron`.
:::

Verify the installation:

```{r}
#| eval: false
cmdstan_version()
```

You should see something like `"2.35.0"`. Record this version — you will
want it for the methods section of any paper you write.


## Step 4: Install companion R packages

The rest of the handbook uses these packages for working with and visualising
Stan output. Install them all now:

```{r}
#| eval: false
install.packages(c(
  "posterior",   # working with posterior draws
  "bayesplot",   # MCMC diagnostics and posterior plots
  "loo",         # leave-one-out cross-validation
  "ggplot2",     # all other plots
  "dplyr",       # data wrangling
  "tidyr",       # reshaping
  "patchwork"    # combining plots
))
```


## The anatomy of a Stan program {#sec-anatomy}

Every Stan program is divided into **named blocks**. Not all blocks are
required, but they must appear in this fixed order when present:

```stan
functions {
  // (optional) user-defined functions
}

data {
  // variables read from R
}

transformed data {
  // (optional) deterministic functions of the data
}

parameters {
  // unknowns to be sampled
}

transformed parameters {
  // (optional) deterministic functions of parameters
}

model {
  // priors and likelihood, the log posterior
}

generated quantities {
  // (optional) predictions, log-likelihoods, derived quantities
}
```

In practice, you will use `data`, `parameters`, and `model` in nearly every
program. The other blocks appear as models grow in complexity.

@tbl-blocks summarises each block's role.

| Block | Required | Purpose |
|---|---|---|
| `functions` | No | Define helper functions reused elsewhere |
| `data` | Yes | Declare inputs passed from R |
| `transformed data` | No | Pre-compute fixed quantities from data |
| `parameters` | Yes | Declare the unknowns Stan samples |
| `transformed parameters` | No | Derived quantities that are also saved |
| `model` | Yes | Log-prior + log-likelihood contributions |
| `generated quantities` | No | Posterior predictions, LOO log-likelihoods |

: The seven named blocks of a Stan program. {#tbl-blocks}

::: {.callout-warning}
**Order matters.** Stan enforces that blocks appear in the order shown above.
A common beginner mistake is writing `model` before `parameters`, in such case Stan will refuse to compile the program.
:::


## Your first Stan program

Let us write, compile, and run a complete Stan program. The model is simple: we observe $N$ measurements drawn from a normal distribution, and we want to learn the mean $\mu$ and standard deviation $\sigma$.

$$
y_i \sim \text{Normal}(\mu,\, \sigma), \quad i = 1, \ldots, N
$$

### Write the Stan file

Create a file called `normal_model.stan` in your working directory with the
following content. (In RStudio: *File => New File => Stan File*.)

```stan
// normal_model.stan
// Estimate the mean and SD of a normal distribution.

data {
  int<lower=1> N;      // number of observations; must be at least 1
  vector[N] y;         // the observations
}

parameters {
  real mu;             // population mean (unconstrained)
  real<lower=0> sigma; // population SD (must be positive)
}

model {
  // Priors
  mu ~ normal(0, 10);      // weakly informative
  sigma ~ exponential(1);     // puts mass on moderate values

  // Likelihood
  y ~ normal(mu, sigma);
}

generated quantities {
  // Posterior predictive draw for one new observation
  real y_rep = normal_rng(mu, sigma);
}
```

### Compile the model

In R, point cmdstanr at the file:

```{r}
#| eval: false
library(cmdstanr)

mod <- cmdstan_model("normal_model.stan")
```

Compilation happens once and is cached. You will see a progress message; when
it finishes, `mod` is a `CmdStanModel` object.

```{r}
#| eval: false
mod   # print a summary
```

```
CmdStanModel: normal_model
  stan_file: normal_model.stan
  exe_file:  normal_model
  ...
```

### Prepare the data

Stan's `data` block expects an R list whose names exactly match the variable
names in the block:

```{r}
#| eval: false
set.seed(2026)

# Simulate some data: true mu = 3, true sigma = 1.5
y_obs <- rnorm(n = 50, mean = 3, sd = 1.5)

stan_data <- list(
  N = length(y_obs),
  y = y_obs
)
```

### Sample from the posterior

```{r}
#| eval: false
fit <- mod$sample(
  data    = stan_data,
  seed    = 2025,
  chains  = 4,         # run 4 independent Markov chains
  parallel_chains = 4, # use 4 CPU cores (adjust to your machine)
  iter_warmup   = 1000,
  iter_sampling = 1000,
  refresh = 500        # print progress every 500 iterations
)
```

You will see output like:

```
Running MCMC with 4 parallel chains...
Chain 1 Iteration:    1 / 2000 [  0%]  (Warmup)
Chain 1 Iteration:  500 / 2000 [ 25%]  (Warmup)
Chain 1 Iteration: 1000 / 2000 [ 50%]  (Warmup)
Chain 1 Iteration: 1001 / 2000 [ 50%]  (Sampling)
Chain 1 Iteration: 1500 / 2000 [ 75%]  (Sampling)
Chain 1 Iteration: 2000 / 2000 [100%]  (Sampling)
Chain 1 finished in 0.1 seconds.
...
All 4 chains finished successfully.
Mean chain execution time: 0.1 seconds.
Total execution time: 0.2 seconds.
```

### Read the output

```{r}
#| eval: false
fit$summary()
```

```
# A tibble: 3 × 10
  variable  mean median    sd   mad    q5   q95  rhat ess_bulk ess_tail
  <chr>    <dbl>  <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>    <dbl>    <dbl>
1 lp__     -86.5  -86.2  1.01 0.726 -88.4 -85.5  1.00    1671.    2332.
2 mu         3.06   3.06 0.214 0.212  2.71  3.41  1.00    3682.    2871.
3 sigma      1.50   1.49 0.153 0.152  1.26  1.77  1.00    3745.    2901.
```

Read this table as follows:

- **`mean` / `median`**: posterior mean and median for each parameter. Here
  $\hat{\mu} \approx 3.06$ and $\hat{\sigma} \approx 1.50$ which are close to the true values of 3 and 1.5.
- **`sd`**: posterior standard deviation (the Bayesian analogue of a standard error).
- **`q5` / `q95`**: the 90% credible interval. We are 90% confident that
  $\mu$ lies between 2.71 and 3.41.
- **`rhat`**: the $\hat{R}$ convergence diagnostic. Values close to 1.00
  (below 1.01) indicate that all chains agree, the sampler has converged.
- **`ess_bulk` / `ess_tail`**: effective sample sizes. Values above ~400 are generally adequate; higher is better.

::: {.callout-tip}
**The `lp__` row** is the log posterior density evaluated at each draw. It is
computed automatically and is useful for diagnosing problems, but you do not
interpret it as a parameter.
:::


## A first look at diagnostics

Even at this early stage, it is good practice to look at the trace plots that shows the visual records of each chain's journey through parameter space:

```{r}
#| eval: false
library(bayesplot)

fit$draws() |>
  mcmc_trace(pars = c("mu", "sigma"))
```

Well-mixed chains look like "fuzzy caterpillars": each chain explores the
same region, with no trends, drift, or sticking (@fig-trace-example).

```{r}
#| label: fig-trace-example
#| echo: false
#| fig-cap: >
#|   Schematic trace plots. Left: problematic chains, one chain
#|   stuck at a different value (poor mixing).  Right: healthy chains, four chains overlapping, exploring the same region.
#| fig-width: 8
#| fig-height: 3

library(ggplot2)
library(patchwork)

set.seed(1)
n_iter <- 200
chains <- 4

# Healthy traces
healthy <- do.call(rbind, lapply(seq_len(chains), function(ch) {
  data.frame(
    iter  = seq_len(n_iter),
    value = cumsum(rnorm(n_iter, 0, 0.08)) + rnorm(1, 3, 0.05),
    chain = paste0("chain:", ch)
  )
}))

p_good <- ggplot(healthy, aes(iter, value, colour = chain)) +
  geom_line(alpha = 0.8, linewidth = 0.4) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "Poor mixing (one chain stuck)", x = "Iteration", y = expression(mu),
       colour = NULL) +
  theme(legend.position = "bottom")

# Stuck chain
stuck <- do.call(rbind, lapply(seq_len(chains), function(ch) {
  offset <- if (ch == 2) 1.8 else 0
  data.frame(
    iter  = seq_len(n_iter),
    value = cumsum(rnorm(n_iter, 0, 0.08)) + rnorm(1, 3 + offset, 0.05),
    chain = paste0("chain:", ch)
  )
}))

p_bad <- ggplot(stuck, aes(iter, value, colour = chain)) +
  geom_line(alpha = 0.8, linewidth = 0.4) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "Healthy mixing", x = "Iteration", y = NULL,
       colour = NULL) +
  theme(legend.position = "bottom")

p_good + p_bad
```

We will return to diagnostics in depth in @sec-anatomy and throughout Part V.
For now, if your trace plots look like the left panel and `rhat` is below
1.01, your model has sampled successfully.


## Saving and reloading a fit

Stan fits can take time. Save them so you do not need to refit:

```{r}
#| eval: false
fit$save_object("normal_model_fit.rds")

# In a new R session:
fit <- readRDS("normal_model_fit.rds")
```


## Troubleshooting

### "C++ compiler not found" (Windows)

Rtools is not on the PATH. Run this in R and then restart:

```{r}
#| eval: false
pkgbuild::find_rtools(debug = TRUE)
```

If it cannot find Rtools, reinstall from
<https://cran.r-project.org/bin/windows/Rtools/> and ensure *"Add Rtools
to system PATH"* is checked.

### "Error: 'make' not found" (macOS)

The Xcode Command Line Tools are missing or need updating:

```bash
xcode-select --install
# or, if already installed but broken:
sudo xcode-select --reset
```

### Compilation succeeds but sampling is very slow

Set `parallel_chains` to the number of physical cores on your machine. Find
this in R with:

```{r}
#| eval: false
parallel::detectCores(logical = FALSE)
```

### "Initialization between (-2, 2) failed after 100 attempts"

Stan initialises parameters randomly in the range (-2, 2) on the *unconstrained* scale. This error means the model's log posterior could not be evaluated at any of these starting points. This is usually a sign of a mis-specified model or data-block mismatch. Double-check that your `stan_data` list names exactly match the `data` block variable names, and that dimensions are correct.

### `rhat` is above 1.01 or chains look like the right panel in @fig-trace-example

Convergence failure. Common causes: too few warmup iterations, a poorly
specified model, or high posterior correlation between parameters. Try
`iter_warmup = 2000` first. We discuss deeper fixes in @sec-debugging.


## Summary

- Stan requires a C++ compiler, CmdStan, and the cmdstanr R package.
  Install them once with `install_cmdstan()` and they are ready for all
  future work.
- Every Stan program has up to seven named blocks in a fixed order:
  `functions`, `data`, `transformed data`, `parameters`,
  `transformed parameters`, `model`, `generated quantities`. The
  three essential ones are `data`, `parameters`, and `model`.
- The workflow is: write a `.stan` file → compile with `cmdstan_model()` →
  pass data as a named R list → call `$sample()` → inspect with
  `$summary()`.
- Always check `rhat` (target: < 1.01) and trace plots before interpreting
  results. A fuzzy-caterpillar trace is a healthy trace.


## Exercises

1. **Verify your installation.** Run the following and paste the output into
   a text file to keep as a record of your environment:
   ```{r}
   #| eval: false
   cmdstan_version()
   packageVersion("cmdstanr")
   packageVersion("posterior")
   R.version.string
   ```

2. **Run the normal model.** Copy `normal_model.stan` from this chapter,
   compile it, generate 100 observations from `rnorm(100, mean = 5, sd = 2)`,
   and fit the model. Does `$summary()` recover the true parameters? What
   happens to the width of the credible intervals if you reduce $N$ to 10?

3. **Experiment with priors.** Change the prior on `mu` from
   `normal(0, 10)` to `normal(0, 0.1)` (a very tight prior centred at zero)
   and refit with the same data. How does the posterior for `mu` change?
   What does this tell you about the influence of the prior relative to the
   data?

4. **Inspect the `generated quantities` block.** The model includes
   `y_rep = normal_rng(mu, sigma)`. Extract this quantity with
   `fit$draws("y_rep")` and plot a histogram. What distribution does it
   approximate, and why?


## Further reading

- cmdstanr documentation: <https://mc-stan.org/cmdstanr/>
- Stan installation guide (all platforms):
  <https://mc-stan.org/docs/cmdstan-guide/installation.html>
- @stan2017 — the primary Stan reference; cite this in any publication
  that uses Stan models.
- @betancourt2017conceptual [§2] — brief, accessible description of HMC
  warmup and why it matters for setting `iter_warmup`.
