# One-Way ANOVA in Stan

```{r}
#| label: setup-ch06
#| include: false
source("../R/setup.R")
```

## Overview

One-way ANOVA is one of the most frequently used tools in life-science research: does the outcome differ across treatment groups? Students learn ANOVA as an $F$-test delivered by `aov()`. This chapter shows that ANOVA is not a separate procedure, it is linear regression with a categorical predictor, and it lives in Stan as naturally as the continuous-predictor models of Chapters 4 and 5.

The chapter introduces two parameterisations of the same model, **treatment coding** and **sum-to-zero coding**, explains why both exist, and shows how to translate Stan posterior output back into the quantities an ANOVA table would report. Along the way, we gain tools that classical ANOVA cannot provide: posterior probabilities for pairwise contrasts, honest uncertainty about group means, and a principled way to handle unbalanced designs.

**Learning objectives**:

By the end of this chapter you will be able to:

- Explain why one-way ANOVA is a special case of linear regression
  and write both as a Stan model.
- Choose between treatment coding and sum-to-zero coding, implement
  each in Stan, and interpret the parameters of each.
- Extract posterior group means, pairwise differences, and their
  credible intervals directly from Stan draws.
- Reconstruct the quantities reported in a classical ANOVA table
  ($F$-statistic, $\eta^2$) from the posterior, and explain their
  limitations.
- Perform a posterior predictive check appropriate for grouped data.


## ANOVA is regression {#sec-anova-is-regression}

Consider a completely randomised experiment with $K$ groups. The classical ANOVA model is:

$$
y_{ij} = \mu + \tau_k + \varepsilon_{ij},
\qquad \varepsilon_{ij} \sim \text{Normal}(0, \sigma)
$$

where $\mu$ is the grand mean, $\tau_k$ is the effect of group $k$, and the constraint $\sum_k \tau_k = 0$ makes the model identifiable.

Now write this as a regression. Create $K - 1$ binary dummy variables $d_{ik}$: $d_{ik} = 1$ if observation $i$ belongs to group $k$, zero otherwise. Then:

$$
y_i = \alpha + \beta_2 d_{i2} + \beta_3 d_{i3} + \cdots + \beta_K d_{iK}
+ \varepsilon_i
$$

Here $\alpha$ is the mean of group 1 (the reference), and $\beta_k$ is the difference between group $k$ and group 1. This is exactly what `lm(y ~ factor(group))` does in R, the ANOVA and the regression are
the same model, just parameterised differently.

In Stan, we can use *either* parameterisation. The choice affects how we interpret the parameters but not the predictions or the fit.

@fig-anova-coding illustrates the two parameterisations side by side.

```{r}
#| label: fig-anova-coding
#| echo: false
#| fig-cap: >
#|   Two parameterisations of the same one-way ANOVA model with four
#|   groups. Treatment coding (left) sets group 1 as the reference; all
#|   parameters are differences from group 1. Sum-to-zero coding (right)
#|   centres all effects around the grand mean; the effects sum to zero
#|   and the last is derived from the others.
#| fig-width: 8
#| fig-height: 3.8
#| code-fold: true
#| cache: true

library(ggplot2)
library(patchwork)

K <- 4
means <- c(1.2, 1.6, 1.0, 1.8)
gname <- paste0("Group ", 1:K)

# Treatment coding diagram
tc_df <- data.frame(
  group = gname,
  mean = means,
  label = c(
    "α\n(ref mean)",
    "α + τ₂",
    "α + τ₃",
    "α + τ₄"
  )
)

p_tc <- ggplot(tc_df, aes(group, mean)) +
  geom_col(fill = "#aec6c8", colour = "#2e7d6a", width = 0.55) +
  geom_hline(yintercept = means[1], linetype = "dashed", colour = "#2e7d6a", linewidth = 0.7) +
  geom_text(aes(label = label), vjust = -0.4, size = 3.2, colour = "#1a2e28") +
  annotate("text", x = 0.55, y = means[1] + 0.06, label = "α", colour = "#2e7d6a", size = 3.5, fontface = "bold") +
  scale_y_continuous(limits = c(0, 2.2)) +
  labs(title = "Treatment coding", x = NULL, y = "Group mean") +
  theme(plot.title = element_text(size = 11))

# Sum-to-zero coding diagram
grand <- mean(means)
tau <- means - grand

sz_df <- data.frame(
  group = gname,
  mean = means,
  label = c(
    sprintf("μ + τ₁\n(%.2f)", tau[1]),
    sprintf("μ + τ₂\n(%.2f)", tau[2]),
    sprintf("μ + τ₃\n(%.2f)", tau[3]),
    sprintf("μ + τ₄*\n(%.2f)", tau[4])
  )
)

p_sz <- ggplot(sz_df, aes(group, mean)) +
  geom_col(fill = "#5c8a52", colour = "#3a6035", alpha = 0.75, width = 0.55) +
  geom_hline(yintercept = grand, linetype = "dashed", colour = "#3a6035", linewidth = 0.7) +
  geom_text(aes(label = label), vjust = -0.3, size = 3.0, colour = "#1a2e28") +
  annotate("text", x = 0.55, y = grand + 0.06,
           label = "μ (grand mean)", colour = "#3a6035",
           size = 3.2, fontface = "bold") +
  annotate("text", x = 4.45, y = 1.55,
           label = "* derived:\nτ₄ = −Στ₁₋₃",
           colour = "#c0392b", size = 2.8, hjust = 1) +
  scale_y_continuous(limits = c(0, 2.2)) +
  labs(title = "Sum-to-zero coding", x = NULL, y = NULL) +
  theme(plot.title = element_text(size = 11))

p_tc + p_sz
```


## Running example: soil carbon under land use types

We model topsoil organic carbon (SOC, g kg⁻¹) across four land use types common in the West African forest-savanna mosaic: natural forest, agroforest, fallow, and cropland. The scientific question is whether land use type is associated with different SOC levels, a one-way ANOVA with $K = 4$ groups.

```{r}
#| label: simulate-data-ch06
#| fig-cap: >
#|   Simulated topsoil organic carbon (SOC, g kg⁻¹) by land use type.
#|   Points are individual plot measurements (jittered horizontally);
#|   horizontal bars are group means. Cropland has the lowest SOC;
#|   natural forest the highest.
#| fig-width: 6.5
#| fig-height: 4
#| code-fold: true
#| cache: true

set.seed(2025)

# True group means (g/kg SOC, broadly consistent with West African
# forest-savanna literature)
K <- 4
group_names <- c("Natural forest", "Agroforest", "Fallow", "Cropland")
true_mu <- c(18.5, 14.2, 11.8, 8.4)   # group means
true_sigma <- 2.8                     # common within-group SD

# Unbalanced design (realistic: fewer forest plots)
n_per_group <- c(10, 14, 14, 12)
N <- sum(n_per_group)

group <- rep(seq_len(K), times = n_per_group)
y <- rnorm(N, mean = true_mu[group], sd = true_sigma)

soc_df <- data.frame(
  y = y,
  group = group,
  group_name = factor(group_names[group], levels = group_names)
)

# Group means for plot
group_means <- tapply(y, soc_df$group_name, mean)
gm_df <- data.frame(
  group_name = names(group_means),
  mean = as.numeric(group_means)
)

ggplot(soc_df, aes(group_name, y)) +
  geom_jitter(width = 0.12, colour = "#2e7d6a", alpha = 0.7, size = 2) +
  geom_crossbar(data = gm_df,
                aes(y = mean, ymin = mean, ymax = mean),
                colour = "#1a4a40", linewidth = 0.9, width = 0.45) +
  labs(
    x = "Land use type",
    y = "Topsoil SOC (g kg⁻¹)",
    title = "Soil organic carbon by land use type"
  )
```

::: {.callout-note}
**Why an unbalanced design?**  
Natural forest plots are often scarcer and harder to access than cropland plots. With $n = (10, 14, 14, 12)$, this is a mildly unbalanced design. Classical ANOVA handles unbalanced designs awkwardly (Type I vs Type III sums of squares). Bayesian ANOVA handles them transparently, the  likelihood simply reflects the different amounts of information in each group.
:::

## Parameterisation 1: treatment coding {#sec-treatment}

### The Stan model

```stan
data {
  int<lower=1> N;
  int<lower=2> K;
  array[N] int<lower=1, upper=K> group;
  vector[N] y;
}

parameters {
  real alpha;         // reference group mean (group 1)
  vector[K-1] tau;    // deviations from reference (groups 2..K)
  real<lower=0> sigma;
}

transformed parameters {
  vector[K] mu;
  mu[1] = alpha;
  for (k in 2:K)
    mu[k] = alpha + tau[k-1];
}

model {
  alpha ~ normal(0, 10);
  tau ~ normal(0, 5);
  sigma ~ exponential(1);

  for (i in 1:N)
    y[i] ~ normal(mu[group[i]], sigma);
}

generated quantities {
  vector[N] y_rep;
  vector[N] log_lik;

  for (i in 1:N) {
    y_rep[i] = normal_rng(mu[group[i]], sigma);
    log_lik[i] = normal_lpdf(y[i] | mu[group[i]], sigma);
  }
}
```

### Key design decisions

**`array[N] int group`** is Stan's syntax for an integer array. One group
index per observation. Stan 2.26+ uses the `array[N] int` declaration;
older syntax was `int group[N]`. Always use the modern form.

**`transformed parameters` block.** We assemble the full group mean vector `mu` here rather than in the `model` block. This means `mu` is saved in the posterior draws automatically, we can inspect each group mean without any post-processing arithmetic.

**`vector[K-1] tau`** declares $K - 1$ free treatment effects. Group 1 is the reference, so `tau[1]` is the group 2 vs group 1 difference, `tau[2]` is group 3 vs group 1, and so on. This is exactly what `lm(y ~ factor(group))` returns.

### Compile and sample

```{r}
#| label: compile-tc
#| eval: false
mod_tc <- cmdstan_model("stan/anova_treatment.stan")
```

```{r}
#| label: compile-tc-real
#| include: false
mod_tc <- cmdstan_model("../stan/anova_treatment.stan")
```

```{r}
#| label: data-list-ch06
stan_data_anova <- list(
  N = N,
  K = K,
  group = group,
  y = y
)
```

```{r}
#| label: sample-tc
#| eval: false
fit_path_tc <- "anova_treatment_fit.rds"

if (!file.exists(fit_path_tc)) {
  fit_tc <- mod_tc$sample(
    data = stan_data_anova,
    seed = 2025,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    refresh = 0
  )
  fit_tc$save_object(fit_path_tc)
} else {
  fit_tc <- readRDS(fit_path_tc)
}
```

```{r}
#| label: sample-tc-real
#| include: false
fit_path_tc <- file.path(
  dirname(knitr::current_input(dir = TRUE)),
  "anova_treatment_fit.rds"
)

if (!file.exists(fit_path_tc)) {
  fit_tc <- mod_tc$sample(
    data = stan_data_anova,
    seed = 2025,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    refresh = 0
  )
  fit_tc$save_object(fit_path_tc)
} else {
  fit_tc <- readRDS(fit_path_tc)
}
```

### Convergence and summary

```{r}
#| label: summary-tc
fit_tc$summary(variables = c("alpha", "tau[1]", "tau[2]",
                              "tau[3]", "sigma",
                              "mu[1]", "mu[2]", "mu[3]", "mu[4]"))
```

```{r}
#| label: fig-trace-tc
#| fig-cap: >
#|   Trace plots for the treatment-coded ANOVA. All chains mix well.
#|   The tau parameters are treatment contrasts vs the reference group
#|   (natural forest); mu parameters are the derived group means.
#| fig-width: 9
#| fig-height: 5
#| code-fold: true
#| cache: true
mcmc_trace(
  fit_tc$draws(c("alpha", "tau[1]", "tau[2]", "tau[3]", "sigma"))
) +
  labs(title = "Trace plots: treatment coding")
```

## Parameterisation 2: sum-to-zero coding {#sec-sumtozero}

### Why sum-to-zero?

Treatment coding makes group 1 special, all other effects are relative to it. The choice of reference group affects the values (and priors) of every `tau` parameter, which can be awkward when no group is a natural reference.

Sum-to-zero coding centres all effects around the grand mean $\mu$:

$$
y_i \sim \text{Normal}(\mu + \tau_{g(i)},\; \sigma),
\qquad \sum_{k=1}^K \tau_k = 0
$$

Now $\mu$ is the grand (unweighted) mean and $\tau_k$ is the deviation of group $k$ from it. The constraint $\sum \tau_k = 0$ is imposed by deriving the last effect: $\tau_K = -\sum_{k=1}^{K-1} \tau_k$.

### The Stan model

```stan
data {
  int<lower=1> N;
  int<lower=2> K;
  array[N] int<lower=1, upper=K> group;
  vector[N] y;
}

parameters {
  real mu;              // grand mean
  vector[K-1] tau_raw;  // free effects for groups 1..K-1
  real<lower=0> sigma;
}

transformed parameters {
  vector[K] tau;
  tau[1:(K-1)] = tau_raw;
  tau[K] = -sum(tau_raw);   // sum-to-zero constraint
}

model {
  mu ~ normal(0, 10);
  tau_raw ~ normal(0, 5);
  sigma ~ exponential(1);

  for (i in 1:N)
    y[i] ~ normal(mu + tau[group[i]], sigma);
}

generated quantities {
  vector[N] y_rep;
  vector[N] log_lik;

  for (i in 1:N) {
    y_rep[i] = normal_rng(mu + tau[group[i]], sigma);
    log_lik[i] = normal_lpdf(y[i] | mu + tau[group[i]], sigma);
  }
}
```

::: {.callout-tip}
**Implementing the sum-to-zero constraint.**  
Stan cannot impose `sum(tau) == 0` as a hard constraint in the `parameters` block — that would reduce the parameter space dimension. Instead, we declare `K-1` free parameters (`tau_raw`) and derive the $K$-th in `transformed parameters` as `tau[K] = -sum(tau_raw)`. Stan samples freely in the $(K-1)$-dimensional unconstrained space and the constraint holds exactly for every draw. This is the standard Stan idiom for sum-to-zero constraints.
:::

### Compile and sample

```{r}
#| label: compile-sz-real
#| include: false
mod_sz <- cmdstan_model("../stan/anova_sumtozero.stan")
```

```{r}
#| label: sample-sz-real
#| include: false
fit_path_sz <- file.path(
  dirname(knitr::current_input(dir = TRUE)),
  "anova_sumtozero_fit.rds"
)

if (!file.exists(fit_path_sz)) {
  fit_sz <- mod_sz$sample(
    data = stan_data_anova,
    seed = 2025,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    refresh = 0
  )
  fit_sz$save_object(fit_path_sz)
} else {
  fit_sz <- readRDS(fit_path_sz)
}
```

```{r}
#| label: summary-sz
fit_sz$summary(variables = c("mu", "tau[1]", "tau[2]",
                              "tau[3]", "tau[4]", "sigma"))
```

::: {.callout-note}
**Both models give identical group means.**  
Treatment coding: group mean $k$ = `alpha + tau[k-1]`. Sum-to-zero coding: group mean $k$ = `mu + tau[k]`. The two parameterisations are mathematically equivalent, they differ only in what the named parameters represent, not in what the model predicts. Verify this by comparing `mu[k]` from the treatment model with `mu + tau[k]` from the sum-to-zero model: they should match to numerical precision.
:::


## Group means and pairwise comparisons {#sec-contrasts}

### Posterior group means

Whether you used treatment or sum-to-zero coding, the group means `mu[k]` are the quantities of primary scientific interest.

```{r}
#| label: fig-group-means
#| fig-cap: >
#|   Posterior distributions of group means (SOC, g kg⁻¹) for the four
#|   land use types. Points are posterior medians; thick bars are 50%
#|   credible intervals; thin bars are 90% credible intervals. Red
#|   crosses mark the true data-generating means. The natural forest
#|   mean is estimated most precisely despite the smallest sample
#|   (n = 10), because its SOC values are more tightly clustered.
#| fig-width: 6.5
#| fig-height: 3.8
#| code-fold: true
#| cache: true

draws_tc <- as_draws_df(fit_tc$draws(c("mu[1]", "mu[2]",
                                        "mu[3]", "mu[4]")))
names(draws_tc)[1:4] <- group_names

mu_long <- draws_tc |>
  select(all_of(group_names)) |>
  pivot_longer(everything(),
               names_to  = "group",
               values_to = "value") |>
  group_by(group) |>
  summarise(
    median = median(value),
    lo50 = quantile(value, 0.25),
    hi50 = quantile(value, 0.75),
    lo90 = quantile(value, 0.05),
    hi90 = quantile(value, 0.95),
    .groups = "drop"
  ) |>
  mutate(
    group = factor(group, levels = group_names),
    true_val = true_mu[as.integer(group)]
  )

ggplot(mu_long, aes(y = group)) +
  geom_segment(aes(x = lo90, xend = hi90, yend = group),
               colour = "#2e7d6a", linewidth = 0.8) +
  geom_segment(aes(x = lo50, xend = hi50, yend = group),
               colour = "#2e7d6a", linewidth = 2.5) +
  geom_point(aes(x = median), colour = "#1a4a40", size = 3) +
  geom_point(aes(x = true_val), colour = "#c0392b",
             shape = 4, size = 3, stroke = 1.2) +
  labs(
    x = "Topsoil SOC (g kg⁻¹)",
    y = NULL,
    title = "Posterior group means",
    subtitle = "Median, 50% CI (thick), 90% CI (thin), ✕ true value"
  )
```

### Pairwise contrasts

Classical ANOVA follows up a significant $F$-test with post-hoc tests (Tukey HSD, Bonferroni). In Stan, pairwise contrasts are just arithmetic on the posterior draws, no adjustment for multiple comparisons is needed because we are estimating probabilities, not controlling Type I error rates.

```{r}
#| label: pairwise-contrasts
#| cache: true
#| code-fold: true
draws_mu <- as_draws_df(fit_tc$draws(c("mu[1]", "mu[2]",
                                        "mu[3]", "mu[4]")))
names(draws_mu)[1:4] <- group_names

# All pairwise differences
pairs <- combn(group_names, 2, simplify = FALSE)

contrast_df <- do.call(rbind, lapply(pairs, function(p) {
  diff <- draws_mu[[p[1]]] - draws_mu[[p[2]]]
  data.frame(
    contrast = paste(p[1], "-", p[2]),
    mean = round(mean(diff), 2),
    lo90 = round(quantile(diff, 0.05), 2),
    hi90 = round(quantile(diff, 0.95), 2),
    p_positive = round(mean(diff > 0), 3)
  )
}))

knitr::kable(
  contrast_df,
  col.names = c("Contrast", "Mean diff.", "5%", "95%",
                "P(diff > 0 | data)"),
  caption = "Posterior pairwise SOC differences (g kg⁻¹). The last
               column is the posterior probability that the first group
               has higher SOC than the second."
)
```

```{r}
#| label: fig-contrasts
#| fig-cap: >
#|   Posterior distributions of all six pairwise SOC differences.
#|   Intervals that exclude zero (vertical dashed line) indicate
#|   clear directional evidence. The natural forest vs cropland
#|   contrast shows the strongest and most certain difference.
#| fig-width: 7
#| fig-height: 4.5
#| code-fold: true
#| cache: true

contrast_draws <- do.call(rbind, lapply(pairs, function(p) {
  diff <- draws_mu[[p[1]]] - draws_mu[[p[2]]]
  data.frame(
    contrast = paste(p[1], "-", p[2]),
    value = diff
  )
}))

contrast_draws |>
  group_by(contrast) |>
  summarise(
    median = median(value),
    lo90 = quantile(value, 0.05),
    hi90 = quantile(value, 0.95),
    lo50 = quantile(value, 0.25),
    hi50 = quantile(value, 0.75),
    .groups = "drop"
  ) |>
  mutate(
    contrast = reorder(contrast, median)
  ) |>
  ggplot(aes(y = contrast)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  geom_segment(aes(x = lo90, xend = hi90, yend = contrast),
               colour = "#2e7d6a", linewidth = 0.8) +
  geom_segment(aes(x = lo50, xend = hi50, yend = contrast),
               colour = "#2e7d6a", linewidth = 2.5) +
  geom_point(aes(x = median), colour = "#1a4a40", size = 3) +
  labs(
    x = "Difference in SOC (g kg⁻¹)",
    y = NULL,
    title = "Posterior pairwise contrasts",
    subtitle = "Median, 50% CI (thick), 90% CI (thin)"
  )
```

## Connecting to the classical ANOVA table {#sec-anova-table}

Students expect to see an ANOVA table with $SS$, $df$, $MS$, and the $F$-statistic. We can reconstruct these from the posterior, but more importantly we can explain *what they actually measure*, and why the Bayesian alternatives are more useful.

### The classical table

```{r}
#| label: classical-anova
aov_fit <- aov(y ~ factor(group), data = soc_df)
summary(aov_fit)
```

### Posterior $R^2$ and $\eta^2$

The classical effect size $\eta^2 = SS_{\text{between}} / SS_{\text{total}}$ measures the proportion of variance explained by group membership. We can compute a Bayesian analogue directly from the posterior.

```{r}
#| label: bayes-r2
#| cache: true
#| code-fold: true
# Bayesian R²
# R² = Var(fitted) / (Var(fitted) + sigma²)
draws_all <- as_draws_df(
  fit_tc$draws(c("mu[1]", "mu[2]", "mu[3]", "mu[4]", "sigma"))
)

# Column names are "mu[1]"..."mu[4]" and "sigma", use them directly
mu_cols <- paste0("mu[", 1:K, "]")

bayes_r2 <- sapply(seq_len(nrow(draws_all)), function(i) {
  # Map each observation to its group mean draw
  mu_obs <- as.numeric(draws_all[i, mu_cols])[group]
  var_fit <- var(mu_obs)
  sigma_i <- draws_all$sigma[i]
  var_fit / (var_fit + sigma_i^2)
})

cat("Posterior mean Bayesian R²: ", round(mean(bayes_r2), 3), "\n")
cat("90% credible interval: [",
    round(quantile(bayes_r2, 0.05), 3), ",",
    round(quantile(bayes_r2, 0.95), 3), "]\n")
cat("Classical eta²:             ",
    round(summary(aov_fit)[[1]]["factor(group)", "Sum Sq"] /
          sum(summary(aov_fit)[[1]][, "Sum Sq"]), 3), "\n")
```

::: {.callout-note}
**Why Bayesian $R^2$ instead of $\eta^2$?**  
Classical $\eta^2$ is a single number computed from point estimates. Bayesian $R^2$ [@gelman2019r2] is a full posterior distribution: it tells you not just the most likely proportion of variance explained, but how uncertain that proportion is. With small samples, the credible interval for $R^2$ can be very wide, a fact $\eta^2$ conceals entirely.
:::

### Reconstructing the $F$-statistic

The $F$-statistic can be recovered from the posterior mean estimates, but this is rarely useful that is the posterior gives richer information directly. We show it here purely for comparison.

```{r}
#| label: f-from-posterior
#| code-fold: true
#| cache: true
# Posterior mean group means
mu_hat  <- colMeans(draws_all[, mu_cols])
sigma_hat <- mean(draws_all$sigma)

# Between-group variance (weighted by n)
grand_hat <- weighted.mean(mu_hat, n_per_group)
ss_between <- sum(n_per_group * (mu_hat - grand_hat)^2)
ms_between <- ss_between / (K - 1)
ms_within  <- sigma_hat^2

cat("F from posterior means:  ", round(ms_between / ms_within, 2), "\n")
cat("F from classical ANOVA:  ",
    round(summary(aov_fit)[[1]]["factor(group)", "F value"], 2), "\n")
```


## Posterior predictive check

```{r}
#| label: fig-ppc-ch06
#| fig-cap: >
#|   Posterior predictive check for the treatment-coded ANOVA.
#|   Observed SOC distributions per group (dark lines) are well
#|   covered by the 50 replicated datasets (light lines). The
#|   wider spread for natural forest (smallest n) is correctly
#|   captured by the model.
#| fig-width: 8
#| fig-height: 4.5
#| cache: true
#| code-fold: true

y_rep_tc <- as_draws_matrix(fit_tc$draws("y_rep"))

ppc_dens_overlay_grouped(
  y = y,
  yrep = y_rep_tc[1:50, ],
  group = soc_df$group_name
) +
  labs(
    title = "Posterior predictive check by land use type",
    subtitle = "Dark: observed. Light: 50 posterior predictive datasets"
  ) +
  theme(legend.position = "none")
```


## Comparison with `aov()` and `lm()`

```{r}
#| label: compare-classical
#| cache: true
#| code-fold: true
# lm() with treatment coding (same parameterisation as Stan model 1)
lm_tc <- lm(y ~ factor(group), data = soc_df)

# Stan posterior means for group means
stan_mu <- fit_tc$summary(paste0("mu[", 1:K, "]")) |>
  select(variable, mean, q5, q95)

# lm() group means (alpha + tau)
lm_mu <- c(
  coef(lm_tc)[1],
  coef(lm_tc)[1] + coef(lm_tc)[-1]
)
lm_ci <- predict(lm_tc,
                  newdata = data.frame(group = factor(1:K)),
                  interval = "confidence", level = 0.90)

comparison <- data.frame(
  group = group_names,
  stan_mean = round(stan_mu$mean, 2),
  stan_lo = round(stan_mu$q5, 2),
  stan_hi = round(stan_mu$q95, 2),
  lm_mean = round(lm_mu, 2),
  lm_lo = round(lm_ci[, "lwr"], 2),
  lm_hi = round(lm_ci[, "upr"], 2),
  true = true_mu
)

knitr::kable(
  comparison,
  col.names = c("Group", "Bayes mean", "Bayes 5%", "Bayes 95%",
                "OLS mean", "OLS 5%", "OLS 95%", "True"),
  caption = "Posterior group means vs OLS estimates. Values are
               nearly identical with weakly informative priors
               and n ≥ 10 per group."
)
```


## Summary

- One-way ANOVA is linear regression with a categorical predictor. In
  Stan, the two natural parameterisations are **treatment coding**
  (effects relative to a reference group) and **sum-to-zero coding**
  (effects relative to the grand mean). Both give identical group means
  and fits.
- Treatment coding uses `real alpha` and `vector[K-1] tau`; the full
  group mean vector is assembled in `transformed parameters`. Sum-to-zero
  coding uses `real mu` and `vector[K-1] tau_raw`, deriving the last
  effect as `tau[K] = -sum(tau_raw)`.
- **Pairwise contrasts** are arithmetic on posterior draws, no
  multiplicity adjustment is required, and the result is a full
  posterior distribution for each difference, not just a $p$-value.
- Bayesian $R^2$ generalises $\eta^2$ to a posterior distribution that
  honestly represents uncertainty about the proportion of variance
  explained.
- Unbalanced designs are handled transparently by the likelihood;
  no special treatment is required.
- The `ppc_dens_overlay_grouped()` function from `bayesplot` provides
  a group-specific predictive check, more informative than the
  marginal overlay used in earlier chapters.


## Exercises

1. **Change the reference group.** Refit the treatment-coded model
   setting cropland (group 4) as the reference. How do `alpha` and the
   `tau` values change? Do the group means `mu[k]` change? Why or why not?

2. **Verify the equivalence.** Using posterior draws from both the
   treatment-coded and sum-to-zero models, compute the group means
   implied by each (`alpha + tau[k-1]` vs `mu + tau[k]`). Show that
   they match to numerical precision using `all.equal()`.

3. **Ordered contrasts.** Ecologists often expect a monotone gradient:
   forest > agroforest > fallow > cropland. Using the pairwise contrast
   draws, compute the posterior probability that *all three* adjacent
   pairs are in the expected direction simultaneously:
   ```{r}
   #| eval: false
   mean(
     (draws_mu[["Natural forest"]] > draws_mu[["Agroforest"]]) &
     (draws_mu[["Agroforest"]] > draws_mu[["Fallow"]]) &
     (draws_mu[["Fallow"]] > draws_mu[["Cropland"]])
   )
   ```
   What is this probability? Is it larger or smaller than you expected?

4. **Heteroscedastic ANOVA.** The homoscedastic assumption (common
   $\sigma$ across groups) can be relaxed. Modify the treatment-coded
   Stan model to estimate a separate `sigma[k]` for each group:
   replace `real<lower=0> sigma` with `vector<lower=0>[K] sigma` and
   update the likelihood to `normal(mu[group[i]], sigma[group[i]])`.
   Refit and compare LOO scores. Is the heteroscedastic model preferred?

5. **Sample size sensitivity.** Rerun the simulation with
   `n_per_group <- c(3, 4, 4, 3)` (very small groups). How do the
   posterior group means compare to the OLS means? Where does the
   Bayesian approach benefit most from the prior?


## Further reading

- @gelman2020regression [Ch. 16]: ANOVA from a regression perspective,
  with a thorough treatment of parameterisation choices.
- @mcelreath2020rethinking [Ch. 5]: categorical predictors and contrasts
  built from posterior draws, with the DAG perspective on group structure.
- @gelman2013bda [Ch. 15]: hierarchical ANOVA as a Bayesian model;
  previews the partial-pooling approach of Chapter 8.
- @gelman2019r2: the reference for Bayesian $R^2$ as a posterior
  distribution rather than a point estimate.
- Stan User's Guide, "ANOVA Models":
  <https://mc-stan.org/docs/stan-users-guide/regression.html>